### 2015-06-17 doc/thesis/system.tex
Back porting the section about types from paper to thesis

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; T_{1} \cup T_{2} & \textit{disjoint union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.T & \textit{recursive types}\\
& | \; \phi(T_{1},T_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; T \; | \; \Const \; T &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
& \;\; \Void & \textit{void type}\\
& | \; T{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types,
filter types, and projection types.
Second-level types include tuple types and unions of tuple types.
Tuple types include the type $\Void$, variadic types, and pair types.
Types are ordered by a subtype relationship that we introduce
in Section \ref{sec:subtyping}, so Lua values may belong to
several distinct types.

@@

Union types $T_{1} \cup T_{2}$ represent data types that can hold a value
of two different types.

@@

Second-level types are either tuple types or unions of tuple types.
Tuple types are tuples of first-level types that can end with
either an empty tuple or with a variadic type.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, only appearing on argument passing, multiple returns,
and multiple assignments.
The type $\Void$ is the type of an empty tuple.
A variadic type $T{*}$ represents a sequence of values of type $T \cup \Nil$;
it is the type of a vararg expression.
Second-level types include unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
For clarity, we use the symbol $\sqcup$ to represent the union between
two different tuple types.
Note that $\cup$ represents the union between two first-level types,
while $\sqcup$ represents the union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{tag}$,
where each $K_{i}$ represents the type of a table key,
and each $V_{i}$ represents the type of the value that table keys of type $K_{i}$ map to.
Key types can only be literal types, base types, or the top type.
We made this restriction to the type of the keys because the statistics
that we discussed in Section \ref{sec:statistics} showed that most
of the tables are records, lists, and hashes.
The type $\Value$ is an option when we need a loose table type.
For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
table in which both indices and values can have any type.
Value types can be any first-level type, and can optionally include
the $\Const$ type to denote immutable values.

@@

Recursive types have the form $\mu x.T$,
where $T$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked lists of integers.
In Section \ref{sec:alias} we mentioned that we can use the following
interface declaration as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

>>>>

Typed Lua includes types that can appear in annotations and some
special types that cannot appear in annotations, though they
play special role in type checking some Lua idioms and handling flow typing.
Special types cannot appear in annotations because they do not have
a corresponding concrete syntax, so we also isolated them in the
formalization to make the implementation easier.
First we present the types that compose the type language of our
type system, and then we present the special types, indicating
why they are necessary.

\begin{figure*}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{F{:}V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
& \;\; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure*}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in the
input and output of functions.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, and recursive types.
Second-level types include tuple types and unions of tuple types.
Tuple types include variadic types and pair types.
Types are ordered by a subtype relationship that we introduce
in Section \ref{sec:subtyping}, so Lua values may belong to
several distinct types.

@@

Union types $F_{1} \cup F_{2}$ represent data types that can hold a value
of two different types.

@@

Second-level types are either tuple types or unions of tuple types.
Tuple types are tuples of first-level types that end with a variadic type.
Typed Lua needs second-level types because tuples are not first-class
values in Lua, only appearing on argument passing, multiple returns,
and multiple assignments.
A variadic type $F{*}$ represents a sequence of values of type $F \cup \Nil$;
it is the type of a vararg expression.
Second-level types include unions of tuples because Lua programs
usually overload the return type of functions to denote error,
as we mentioned in Section \ref{sec:statistics}.
For clarity, we use the symbol $\sqcup$ to represent the union between
two different tuple types.
Note that $\cup$ represents the union between two first-level types,
while $\sqcup$ represents the union between two tuple types.

Back to first-level types, table types represent the various forms
that Lua tables can take.
The syntactical form of table types is $\{\overline{F{:}V}\}_{unique|open|fixed|closed}$,
where the notation $\overline{F{:}V}$ denotes the list $F_{1}{:}V_{1}, ..., F_{n}{:}V_{n}$.
Each $V_{i}$ represents the type of the value that table keys of type $F_{i}$ map to.
%Key types can only be literal types, base types, or the top type.
%We made this restriction to the type of the keys because the statistics
%that we discussed in Section \ref{sec:statistics} showed that most
%of the tables are records, lists, and hashes.
%The type $\Value$ is an option when we need a loose table type.
%For instance, $\{\Value:\Value\}_{closed}$ represents the type of a
%table in which both indices and values can have any type.
Value types represent mutable values by default, but we can use the
$\Const$ type to make them represent immutable values.

@@

Recursive types have the form $\mu x.F$,
where $F$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked lists of integers.
In Section \ref{sec:alias} we mentioned that we can use the following
interface declaration as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

With the recursive types we finish the dicussion about Typed Lua types,
and we begin the discussion about the special types.

\begin{figure*}[!ht]
\textbf{Special Types}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \textsc{expression types:}\\
& \;\; F & \textit{first-level types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
%\multicolumn{3}{c}{}\\
E ::= & & \textsc{expression list types:}\\
& \;\; T{*} & \textit{variadic types}\\
& | \; T \times E & \textit{pair types}\\
%\multicolumn{3}{c}{}\\
R ::= & & \textsc{filter result types:}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F & \textit{first-level types}
\end{array}
$$
\dend
\caption{The special types used by Typed Lua}
\label{fig:specialtypes}
\end{figure*}

Figure \ref{fig:specialtypes} presents the special types that
Typed Lua includes for typing some Lua idioms and flow typing.
Typed Lua splits special types into three categories:
\emph{expression types}, \emph{expression list types}, and
\emph{filter result types}.
Expression types represent the type of expressions in the type system,
which can be any first-level type, a filter type, or a projection type.
Expression list types represent the type of lists of expressions,
which are tuples of expression types that end with a variadic type.
Filter result types are either the type $\Void$ or a first-level type.
In Section \ref{sec:fap} we will show that Typed Lua uses filter types
in flow typing, being the type $\Void$ a type with no values, used by the
type system as a way to detect branches that are unreachable due to flow typing.
